---
title: "Fitting a Poisson-lognormal distribution in Stan (rstan)"
author: "Masatoshi Katabuchi"
date: "September 2018"
output:
  #rmdformats::material
  html_document:
    fig_caption: yes
    theme: spacelab #readable #sandstone #spacelab #flatly
    highlight: pygments #tango #kate
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
#library(rmdformats)
```

<a href="https://twitter.com/mattocci"><i class="fa fa-twitter fa-1x"></i> Twittter</a>
<a href="https://github.com/mattocci27/"><i class="fa fa-github fa-1x"></i> Github</a>
<a href="mailto:mattocci27@gmail.com"><i class="fa fa-envelope fa-1x"></i> Email</a>

# Poisson-lognormal distribution

A compound Poisson-lognormal distribution is a Poisson probability distribution where its parameter $\lambda$ is a random variable with lognormal distribution, that is to say log$\lambda$ are normally distributed with mean $\mu$ and variance $\sigma^2$ (Bulmer 1974). The density function is

$$
P(k) = \frac{1}{\sqrt{2\pi\sigma^2}k!}\int^\infty_0\lambda^{k-1}exp(-\lambda)exp(\frac{-(log\lambda-\mu)^2}{2\sigma^2})d\lambda, \; \text{where} \; k = 0, 1, 2, ... \;.
$$

The likelihood of the Poisson-lognormal distribution needs to be evaluated by numerical integration. Writing its log-likelihood function from scratch is a pain. In Stan, we can define the likelihood of the Poisson-lognormal distribution using priors literally (i.e., log$\lambda$ are normally distributed with mean $\mu$ and variance $\sigma^2$) as follows;

$$
\sigma \sim \text{Half-}Cauchy(0, 5)
$$
$$
\mu \sim N(0, 10)
$$
$$
log\lambda_i \sim N(\mu, \sigma)
$$
$$
y_i \sim Pois(\lambda_i) \; \text{or} \; y_i \sim PoisLog(log\lambda_i)
$$

where $y_i$ is given as data and $PoisLog$ is log parameterization of the Poisson (it's different from a compound Poisson-lognormal distribution) (Stan Version 2.17.0 Manual page 520).


```{r}
writeLines(readLines("poilog.stan"))
```

# Testing

We generate dummy data, a sample of 100 random positive integers drawn from a Poisson distribution with mean of log$\lambda_i$ which are normally distributed with a mean of log(50) $\simeq$ 3.91 and a standard deviation of 1.0.

```{r}
set.seed(123)
N <- 100
mu <- log(50)
sigma <- 1
log_lambda <- rnorm(N, mu, sigma)
Y <- rpois(N, exp(log_lambda))
```

```{r, cache=TRUE}
library(ggthemes)
theme_set(theme_solarized(light=FALSE))
dat <- data_frame(Y)

ggplot(dat, aes(Y)) +
  geom_histogram(bins=30, fill = "#268bd2", col = "black")

```

We then use the Stan to fit the data.

```{r, cache=TRUE, comment=NA}

library(rstan)
library(loo)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

list_dat <- list(N = N, y = Y)

fit <- stan(file = "poilog.stan",
          data = list_dat,
          iter = 1000,
          warmup = 500,
          thin = 1,
          chains =  4,
          refresh = 500,
          control = list(adapt_delta = 0.9, max_treedepth = 20))

```

```
SAMPLING FOR MODEL 'poilog' NOW (CHAIN 1).

Gradient evaluation took 6.4e-05 seconds
1000 transitions using 10 leapfrog steps per transition would take 0.64 seconds.
Adjust your expectations accordingly!


Iteration:   1 / 1000 [  0%]  (Warmup)
Iteration: 500 / 1000 [ 50%]  (Warmup)
Iteration: 501 / 1000 [ 50%]  (Sampling)
Iteration: 1000 / 1000 [100%]  (Sampling)

 Elapsed Time: 0.262078 seconds (Warm-up)
               0.187632 seconds (Sampling)
               0.44971 seconds (Total)


SAMPLING FOR MODEL 'poilog' NOW (CHAIN 2).

Gradient evaluation took 4.5e-05 seconds
1000 transitions using 10 leapfrog steps per transition would take 0.45 seconds.
Adjust your expectations accordingly!


Iteration:   1 / 1000 [  0%]  (Warmup)
Iteration: 500 / 1000 [ 50%]  (Warmup)
Iteration: 501 / 1000 [ 50%]  (Sampling)
Iteration: 1000 / 1000 [100%]  (Sampling)

 Elapsed Time: 0.254045 seconds (Warm-up)
               0.1933 seconds (Sampling)
               0.447345 seconds (Total)


SAMPLING FOR MODEL 'poilog' NOW (CHAIN 3).

Gradient evaluation took 3.5e-05 seconds
1000 transitions using 10 leapfrog steps per transition would take 0.35 seconds.
Adjust your expectations accordingly!


Iteration:   1 / 1000 [  0%]  (Warmup)
Iteration: 500 / 1000 [ 50%]  (Warmup)
Iteration: 501 / 1000 [ 50%]  (Sampling)
Iteration: 1000 / 1000 [100%]  (Sampling)

 Elapsed Time: 0.229415 seconds (Warm-up)
               0.185794 seconds (Sampling)
               0.415209 seconds (Total)


SAMPLING FOR MODEL 'poilog' NOW (CHAIN 4).

Gradient evaluation took 4.2e-05 seconds
1000 transitions using 10 leapfrog steps per transition would take 0.42 seconds.
Adjust your expectations accordingly!


Iteration:   1 / 1000 [  0%]  (Warmup)
Iteration: 500 / 1000 [ 50%]  (Warmup)
Iteration: 501 / 1000 [ 50%]  (Sampling)
Iteration: 1000 / 1000 [100%]  (Sampling)

 Elapsed Time: 0.230132 seconds (Warm-up)
               0.187833 seconds (Sampling)
               0.417965 seconds (Total)
```

The resulting estimates well recovered the original values ($\mu$ and $\sigma$) `r emo::ji("beer")`


```{r, eval = F, echo = F}
fit
```

```

print(fit)

Inference for Stan model: poilog.
4 chains, each with iter=1000; warmup=500; thin=1;
post-warmup draws per chain=500, total post-warmup draws=2000.

                   mean se_mean   sd    2.5%     25%     50%     75%   97.5% n_eff Rhat
mu                 4.00    0.00 0.10    3.80    3.93    4.00    4.06    4.18  2000 1.00
sigma              0.94    0.00 0.07    0.82    0.89    0.94    0.99    1.10  2000 1.00
log_lambda[1]      3.20    0.00 0.20    2.79    3.06    3.20    3.33    3.56  2000 1.00
log_lambda[2]      3.66    0.00 0.16    3.35    3.56    3.67    3.76    3.95  2000 1.00
log_lambda[3]      5.53    0.00 0.06    5.41    5.49    5.53    5.58    5.65  2000 1.00
.....

log_lambda[97]     6.14    0.00 0.05    6.05    6.11    6.14    6.17    6.24  2000 1.00
log_lambda[98]     5.48    0.00 0.07    5.35    5.43    5.48    5.52    5.61  2000 1.00
log_lambda[99]     3.61    0.00 0.16    3.28    3.50    3.61    3.72    3.91  2000 1.00
log_lambda[100]    2.83    0.01 0.23    2.36    2.67    2.84    3.00    3.25  2000 1.00
lp__            -385.12    0.30 7.16 -400.14 -389.95 -384.88 -380.07 -372.43   580 1.01

Samples were drawn using NUTS(diag_e) at Fri Sep 21 18:17:19 2018.
For each parameter, n_eff is a crude measure of effective sample size,
and Rhat is the potential scale reduction factor on split chains (at
convergence, Rhat=1).

```

# Computing Environment

```{r}
writeLines(readLines(file.path(Sys.getenv("HOME"), ".R/Makevars")))
devtools::session_info("rstan")
```

# Related

- [Poisson lognormal R function](https://www.rdocumentation.org/packages/poilog/versions/0.4/topics/Poisson%20lognormal)

- M. G. Bulmer. 1974. On Fitting the Poisson Lognormal Distribution to Species-Abundance Data. Biometrics 30: 101-110 [doi:10.2307/2529621](https://www.jstor.org/stable/2529621?origin=crossref&seq=1#metadata_info_tab_contents)

- Engen, S., R. Lande, T. Walla & P. J. DeVries. 2002. Analyzing spatial structure of communities using the two-dimensional Poisson lognormal species abundance model. American Naturalist 160: 60-73. [doi:10.1086/340612](https://www.journals.uchicago.edu/doi/abs/10.1086/340612?journalCode=an)
